Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-11 paper.tex
text updated

2. The dwarf spheroidal galaxies orbiting the Milky Way, like Segue-1,
Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

@@

For a satellite dwarf galaxy, however, once it is accreted by the
Milky Way, the outer regions are severely affected by tidal
stripping. The longer a satellite has been part of our Galaxy, and the
closer it comes to the center during its pericentral passage, the more
material is removed \cite{2004MNRAS.355..819G}.

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four promising clusters with a high gamma-ray
yield. We derive the emission from two different DM models and a
hadronic CR model and find that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement is
small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax
($\mvir=10^{14}\,\msun$) and Virgo ($\mvir=2.1\times10^{14}\,\msun$)
\del{\cite{}} as the prime targets for DM observations, Perseus
($\mvir=7.7\times10^{14}\,\msun$) for CR induced emission and we
additionally decide in favor of the well studied cluster Coma
($\mvir=1.4\times10^{15}\,\msun$) for comparison. While studying these
four clusters in great detail, we complement the analysis by
calculating the expected gamma-ray fluxes of a nearby X-ray flux
complete sample of galaxy clusters from the Rosat all-sky survey
(extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} \\
         & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ 
         & & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ \\
 \hline
Fornax               & $4.8\times10^{-8}$ & $1.3\times10^{-10}$ &  $2.2\times10^{-10}$ 
& \,\,\,\,\, & $5.6\times10^{-12}$  & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ \\
Coma                 & $1.3\times10^{-8}$ & $8.3\times10^{-11}$ &  $4.0\times10^{-9}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ \\
Virgo                & $1.2\times10^{-7}$ & $4.0\times10^{-10}$ &  $8.7\times10^{-9}$  
& \,\,\,\,\, & $1.3\times10^{-11}$  & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ \\
Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$ & $7.1\times10^{-11}$ &  $1.4\times10^{-8}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
 (2) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
 model where the boost from substructures is included.\\
 (3) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{Hooper:2010mq}), the dwarf galaxies surrounding the Milky Way
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark LP model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of ($1/2,\,1/2$)
into ($\mu^+/\mu^-,\,\e^+/\e^-$) \footnote{This is an approximation to
  the model found in \cite{Finkbeiner:2010sm} that has a branching
  ratio of ($1/3,\,1/3,\,1/3$) into
  ($\mu^+/\mu^-,\,\e^+/\e^-,\,\pi^+/\pi^-$).}. The Sommerfeld boost
$S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way halo in our LP
model, which we appropriately scale down for cluster halos, using
\del{To connect easily to our previous work
  \cite{2009PhRvL.103r1302P}, we choose as our benchmark LP model one
  with mass $1.6$~TeV and Sommerfeld boost
  $S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way halo and
  scale this appropriately down for cluster halos, using} {\bf Anders:
  insert Sommerfeld boost formula as a function of halo mass, done}
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

\subsubsection{Supersymmetric dark matter}
The most studied models for DM are supersymmetric ones, especially
models where the lightest supersymmetric particle, in most models the
lightest neutralino, is stable. The stability is assured in viable
SUSY models due to a discrete symmetry, $R$-parity. This symmetry is
needed from the particle physics point of view to avoid fast proton
decay, and automatically makes the lightest supersymmetric particle
stable, and therefore a good candidate for DM. In addition, as the
coupling to ordinary matter is given by gauge couplings, the
neutralino is automatically a WIMP candidate. Unfortunately, the
breaking of supersymmetry (which has to be there since otherwise,
e.g., there would exist a scalar electron with the same mass as the
electron) means that a large number of essentially free parameters
enter actual calculations.  The neutralino is then a linear
combination of the supersymmetric partners of the U(1) gauge boson,
$\tilde B$, one of the SU(2) gauge bosons, $\tilde W_3$, and the two
Higgs doublets in the Minimal Supersymmetric extension of the Standard
Model (the MSSM), $\tilde H_1^0$ and $\tilde H_2^0$, with mixing
coefficients which depend on the SUSY breaking parameters
\cite{1984NuPhB.238..453E}. The parameter space is very large, too
large in general to scan efficiently even with the most powerful
computers. Therefore, often simplifying assumptions are made to limit
the number of free parameters. Even then, a large number of viable
models is found which give a relic density consistent with the WMAP
data. This means that they are indeed very good templates for WIMP DM,
something that explains their popularity besides the subjective
statement that few more plausible models for DM have been put forward
so far.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of Milky Way (MW) type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleting their central number densities. Since the
DM annihilation rate depends on the density squared, the resulting
flux from substructures is boosted compared to the smooth density
distribution. Recent high resolution simulations suggest a flux
enhancement from substructures of more than 100
\cite{2008MNRAS.391.1685S, 2010ApJ...718..899A} for a MW size halo. We
adopt a boost from substructures of 230 which is the enhancement found
in one of the highest resolved MW halos ever simulated.\footnote{Note
  that these high values for boost factors are now consistently found
  in high-resolution simulations of dark matter halos
  \cite{2008MNRAS.391.1685S, 2010ApJ...718..899A} after accounting for
  the tidal mass loss of subhalos. This effect causes a significantly
  flatter subhalo distribution compared to the smooth distribution and
  errornously assuming an identical mass density distribution of
  smooth DM and substructure biases the subhalo boost factor too low
  \cite{2008JPhCS.125a2008K, 2008JPhCS.125a2008K}.} We use a double
power-law function to fit \footnote{Our approach of fitting the
  scaling behavior of $L_\rmn{sub}(<r)$ directly from numerical
  simulations self-consistently accounts for the radial dependence of
  the substructure concentration \cite{2008MNRAS.391.1685S}.} the
luminosity from the smooth component of substructures
(i.e. substructures within substructures are not included) inside
radius $r$ which is determined for the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}. Our best fit is given
by
\begin{eqnarray}
  L_\rmn{sub}(<x) &=& a_0\,C(\mvir)\,L_{200\sm}(\mvir)\,x^{a_1
    \displaystyle{x}^{a_2}}\,,\label{eq:LSub}\\
   a_0 &=& 0.76\,,\quad a_1 = 0.95\,,\quad  a_2 = -0.27\,, \quad 
 \nonumber\\
  C(\mvir) &=& \left(\frac{\mvir\,M_{\rm res,sim}}
{M_{\rm 200sim}\,M_{\rm lim}}\right)^{\alpha_C}
=0.023\,\left(\frac{\mvir}{M_{\rm lim}}\right)^{\alpha_C}.\nonumber\\
\label{eq:Csub}
\end{eqnarray}
Here, $\alpha_C=0.226$, $a_i$ denote our fit variables, $L_{200\sm}$
is the luminosity from the smooth halo without substructures within
$\rvir$ and $x= r/\rvir$.  We derive the normalization function
$C(\mvir)$ in Eq.~(\ref{eq:Csub}) from the simulations in reference
\cite{2008Natur.456...73S} using a value of $M_{\rm
  200sim}=1.9\times10^{12}\msun$ for the mass of the MW halo in the
simulation and $M_{\rm res,sim}=10^5\msun$ for the mass of the
smallest resolved subhalos in the MW simulation.  The smallest mass of
subhalos in reality, $M_{\rm lim}$, is determined from the free
streaming length of DM at decoupling---an effect that erases structure
on scales smaller than the free streaming length. In applying
Eq.~(\ref{eq:Csub}), we implicitly assume that the mass power-law
scaling relation is valid down to the free streaming mass of DM
halos. In the CDM universe, this is conventionally taken to be
$10^{-6}\msun$ \cite{2001PhRvD..64h3507H, 2005JCAP...08..003G}. Note
that potentially the power-law could flatten towards smaller mass
scales although current simulations show no hits of such a behavior
and since Einstein's gravity is a scale-free theory, we do not expect
such a behavior on theoretical grounds either.  For DM halos more
(less) massive than the MW we expect a larger (smaller) boost from
substructures, simply because of the larger (smaller) mass range down
to the minimum mass $M_\rmn{lim}$.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters.  In the leptophilic model, DM annihilation radiation
includes final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider a few supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock accelerated CRs.

@@

Because of the large total boost factor for the LP model ($\sim
10^4$), we overproduce the upper gamma-ray limit in the $1-10$~GeV
energy interval set by Fermi-LAT by about a factor 100. This strongly
constrains the boost from both substructures and the Sommerfeld
enhancement. Additionally, these constraints might improve with future
more sensitive Cherenkov telescopes such as CTA. However, considering
that an indirect detection of DM in clusters relies on the boost from
substructures whose main contribution comes from the periphery, we
conclude that the wide angular extent of clusters on the sky in
gamma-rays suggest that these sources are not ideal for IACTs since
their sensitivity drops linearly with source extension.

>>>>

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like
Segue-1, Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

@@

For a satellite dwarf galaxy, however, once it is accreted by the MW,
the outer regions are severely affected by tidal stripping. The longer
a satellite has been part of our Galaxy, and the closer it comes to
the center during its pericentral passage, the more material is
removed \cite{2004MNRAS.355..819G}.

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four promising clusters with a high gamma-ray
yield. We derive the emission from two different DM models and a
hadronic CR model and find that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo \del{\cite{}} as the prime targets for DM observations, Perseus
for CR induced emission and we additionally decide in favor of the
well studied cluster Coma for comparison. While studying these four
clusters in great detail, we complement the analysis by calculating
the expected gamma-ray fluxes of a nearby X-ray flux complete sample
of galaxy clusters from the Rosat all-sky survey (extended HIghest
X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $106$ & $730$  \\
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $44$  & $1310$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $83$  & $870$  \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $54$  & $1280$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement.\\
  (2) The boost from substructures have been intergated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included.\\
  (5) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{Hooper:2010mq}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/v_{\rm
  cl}$.

As LP models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark LP model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of ($1/2,\,1/2$)
into ($\mu^+/\mu^-,\,\e^+/\e^-$) \footnote{This is an approximation to
  the model found in \cite{Finkbeiner:2010sm} that has a branching
  ratio of ($1/3,\,1/3,\,1/3$) into
  ($\mu^+/\mu^-,\,\e^+/\e^-,\,\pi^+/\pi^-$).}. The SFE
$S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a MW halo in our LP model,
which we appropriately scale down for cluster halos, using \del{To
  connect easily to our previous work \cite{2009PhRvL.103r1302P}, we
  choose as our benchmark LP model one with mass $1.6$~TeV and
  Sommerfeld boost $S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a MW
  halo and scale this appropriately down for cluster halos, using}
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

\subsubsection{Supersymmetric dark matter}
The most studied models for DM are supersymmetric ones, especially
models where the lightest supersymmetric particle, in most models the
lightest neutralino, is stable. The stability is assured in viable
supersymmetric models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for DM. In addition,
as the coupling to ordinary matter is given by gauge couplings, the
neutralino is automatically a WIMP candidate. Unfortunately, the
breaking of supersymmetry (which has to be there since otherwise,
e.g., there would exist a scalar electron with the same mass as the
electron) means that a large number of essentially free parameters
enter actual calculations.  The neutralino is then a linear
combination of the supersymmetric partners of the U(1) gauge boson,
$\tilde B$, one of the SU(2) gauge bosons, $\tilde W_3$, and the two
Higgs doublets in the Minimal Supersymmetric extension of the Standard
Model (the MSSM), $\tilde H_1^0$ and $\tilde H_2^0$, with mixing
coefficients which depend on the supersymmetric breaking parameters
\cite{1984NuPhB.238..453E}. The parameter space is very large, too
large in general to scan efficiently even with the most powerful
computers. Therefore, often simplifying assumptions are made to limit
the number of free parameters. Even then, a large number of viable
models is found which give a relic density consistent with the WMAP
data. This means that they are indeed very good templates for WIMP DM,
something that explains their popularity besides the subjective
statement that few more plausible models for DM have been put forward
so far.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of MW type halos find substantial
amount of substructures in the periphery of DM halos, while the
substructures in the center suffer from dynamical friction and tidal
effects depleting their central number densities. Since the DM
annihilation rate depends on the density squared, the resulting flux
from substructures is boosted compared to the smooth density
distribution. Recent high resolution simulations suggest a flux
enhancement from substructures of more than 100
\cite{2008MNRAS.391.1685S, 2010ApJ...718..899A} for a MW size halo. We
adopt a boost from substructures of 230 which is the enhancement found
in one of the highest resolved MW halos ever simulated.\footnote{Note
  that these high values for boost factors are now consistently found
  in high-resolution simulations of dark matter halos
  \cite{2008MNRAS.391.1685S, 2010ApJ...718..899A} after accounting for
  the tidal mass loss of subhalos. This effect causes a significantly
  flatter subhalo distribution compared to the smooth distribution and
  errornously assuming an identical mass density distribution of
  smooth DM and substructure biases the subhalo boost factor too low
  \cite{2008JPhCS.125a2008K, 2008JPhCS.125a2008K}.} We use a double
power-law function to fit \footnote{Our approach of fitting the
  scaling behavior of $L_\rmn{sub}(<r)$ directly from numerical
  simulations self-consistently accounts for the radial dependence of
  the substructure concentration \cite{2008MNRAS.391.1685S}.} the
luminosity from the smooth component of substructures
(i.e. substructures within substructures are not included) inside
radius $r$ which is determined for the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}. Our best fit is given
by
\begin{eqnarray}
  L_\rmn{sub}(<x) &=& a_0\,C(\mvir)\,L_{200\sm}(\mvir)\,x^{a_1
    \displaystyle{x}^{a_2}}\,,\label{eq:LSub}\\
   a_0 &=& 0.76\,,\quad a_1 = 0.95\,,\quad  a_2 = -0.27\,, \quad 
 \nonumber\\
  C(\mvir) &=& \left(\frac{\mvir\,M_{\rm res,sim}}
{M_{\rm 200sim}\,M_{\rm lim}}\right)^{\alpha_C}
=0.023\,\left(\frac{\mvir}{M_{\rm lim}}\right)^{\alpha_C}.\nonumber\\
\label{eq:Csub}
\end{eqnarray}
Here, $\alpha_C=0.226$, $a_i$ denote our fit variables, $L_{200\sm}$
is the luminosity from the smooth halo without substructures within
$\rvir$ and $x= r/\rvir$.  We derive the normalization function
$C(\mvir)$ in Eq.~(\ref{eq:Csub}) from the simulations in reference
\cite{2008Natur.456...73S} using a value of $M_{\rm
  200sim}=1.9\times10^{12}\msun$ for the mass of the MW halo in the
simulation and $M_{\rm res,sim}=10^5\msun$ for the mass of the
smallest resolved subhalos in the MW simulation.  The smallest mass of
subhalos in reality, $M_{\rm lim}$, is determined from the free
streaming length of DM at decoupling---an effect that erases structure
on scales smaller than the free streaming length. In applying
Eq.~(\ref{eq:Csub}), we implicitly assume that the mass power-law
scaling relation is valid down to the free streaming mass of DM
halos. In the CDM universe, this is conventionally taken to be
$10^{-6}\msun$ \cite{2001PhRvD..64h3507H, 2005JCAP...08..003G}. Note
that potentially the power-law could flatten towards smaller mass
scales although current simulations show no hits of such a behavior
and since Einstein's gravity is a scale-free theory, we do not expect
such a behavior on theoretical grounds either.  For DM halos more
(less) massive than the MW we expect a larger (smaller) boost from
substructures, simply because of the larger (smaller) mass range down
to the minimum mass $M_\rmn{lim}$.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters.  In the LP model, DM annihilation radiation includes
final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider a few supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock accelerated CRs.

@@

Because of the large total boost factor for the LP model ($\sim
10^4$), we overproduce the upper gamma-ray limit in the $1-10$~GeV
energy interval set by Fermi-LAT by about a factor 100. This strongly
constrains the boost from both substructures and the
SFE. Additionally, these constraints might improve with future more
sensitive Cherenkov telescopes such as CTA. However, considering that
an indirect detection of DM in clusters relies on the boost from
substructures whose main contribution comes from the periphery, we
conclude that the wide angular extent of clusters on the sky in
gamma-rays suggest that these sources are not ideal for IACTs since
their sensitivity drops linearly with source extension.
